\hl{Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis}

### 2018-04-17 jhir/latex-manuscript/manuscript_highlight.tex
finalized revised manuscript

\usepackage{rotating}
\usepackage{tabularx}
\usepackage{longtable,tabu}
\usepackage[flushleft]{threeparttable}
\usepackage{caption}
\captionsetup{justification=raggedright,singlelinecheck=off,labelfont=bf}
\usepackage{color, colortbl}
\definecolor{lightgraycol}{gray}{0.95}

@@

\title{Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis
}

@@

\begin{abstract}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change. Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research. A significant barrier to this research is the time and resource-intensive cognitive tasks, including the manual coding and sequential analysis of coded transcripts, that are an integral part of the qualitative analysis of patient-counselor communication. Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating these cognitive tasks. In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had been previously annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients' motivational statements to guide clinical practice.

\keywords{hidden markov model \and closed frequent pattern mining \and motivational interviewing \and pediatric obesity}
\end{abstract}

\section{Introduction}
\label{sec:intro}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change~\cite{miller2013motivational}. The theory underlying MI's clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills~\cite{miller2013motivational}. The relational hypothesis suggests that counselors' use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change. MI's technical hypothesis~\cite{miller2009toward} states that counselors' use of communication techniques consistent with the MI framework (``MI-consistent'' or MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''. Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change. Previous studies~\cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention~\cite{walker2011influence}. In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk~\cite{magill2014technical}. These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers' ability to test MI's technical hypothesis.

Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations~\cite{bakeman1997observing, bakeman2011sequential}. Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN. A follow up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}. Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk. Gaume et al.~\cite{gaume2008counsellor} used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service. They found MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed. A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk~\cite{gaume2010counselor}. In this sample, MIIN led to greater sustain talk, but was unrelated to change talk. Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk. Glynn and colleagues~\cite{glynn2014change} linked reflections of change talk to the elicitation of change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use. In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy. A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk. Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in which the various MICO strategies are effective.

The sequential analysis procedure used in the above MI process studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} is based on the first-order Markov Chain model~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}. The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event. Hence, Markov chain models have two main drawbacks. The first is their inability to preserve the long-range dependencies between observations in a sequence. However, in MI, an observed behavior can be influenced by any of the preceding behaviors. The second drawback is their inability to consider similarities between behavior codes and, consequently, their inability to identify multiple similar behaviors that lead to the same outcome. Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences. Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-term dependencies between behaviors, to identify causal relationships. The goal of the current research is to test the applicability of pattern data mining and machine-learning methods to identify effective patterns of patient-counselor communication.

Several prior works have reported the results of adopting machine learning methods, such as topic models~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, classification methods~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} and neural networks~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} to the tasks of annotating MI transcripts and assessing intervention fidelity. Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews. Because MI fidelity is assessed in part by the ratio of counselor questions to reflection (a minimum goal is 1:1), their system employed a SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features achieving nearly 90\% accuracy. In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}. The SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of MI transcripts with 17 behavior codes, which is comparable to human coders. In a follow up study~\cite{hasan2018predicting}, we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk). RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews. This paper represents the next phase in this line of research. We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.

HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden state'' corresponds to a different distribution. Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state. HMMs were originally proposed for speech recognition~\cite{rabiner1989tutorial}, in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}. For example, a HMM-based classifier was applied Doppler ultrasound imaging data to extract features from from the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}. In another study, HMM was used to capture important
characteristics of protein families~\cite{amrhein2003client}. In the application of HMM to patient-counselor communication, hidden states, and the sets of related behavior codes associated with the hidden states, may correspond a patient’s underlying motivational state during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans, and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that are likely to lead to patient change talk and commitment language.

Frequent pattern mining~\cite{aggarwal2014frequent} is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together. Agrawal and Srikant~\cite{agrawal1994fast} first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases. Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, medical imaging~\cite{olukunle2002fast}, chemical and biological analysis~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, web mining~\cite{srivastava2000web}, and outlier analysis~\cite{aggarwal2015outlier}. No published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication. A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns. To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces fewer patterns in more compact form that are easier to interpret. In this study, we leveraged FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk. FPClose is the best closed frequent itemset mining algorithm, which has shown highest performance in terms of running time and memory consumption.

In this paper, we focus on computational methods to facilitate the sequential analysis of MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions. Analysis of these patterns will provide empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge will inform MI theory by providing additional evidence to support MI's technical hypothesis. It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication. This paper is the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions. Bertholet et al.~\cite{bertholet2010change} used HMM to identify hidden states in a brief motivational intervention. Namely, they limited their HMM model to three hidden states which they characterized as ``towards change'', ``away from change'' and ``non-determined''. These states were then used as predictors in a bivariate regression model to predict drinking outcomes 12 months post-intervention. In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication. The goal of this study to evaluate the utility of using HMM and frequent pattern mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions. These two approaches offer the following advantages over the first-order Markov chain based methods most typically used in MI research. First-order Markov chain models identify the likely transitions between individual behaviors. In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications. Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors. Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communications during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.

\section{Materials and Methods}
\label{sec:materialsAndMethods}

\subsection{Dataset}
\label{subsec:dataset}
The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider}. The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational}, and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

\begin{small}
\begin{longtable}{p{1.3cm}p{2cm}p{5cm}p{4.8cm}}
\caption{MYSCOPE codebook} 
\label{tab:codebook} \\
\hline\noalign{\smallskip}
Annotation & Behavior & Description & Example \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
\endfirsthead
\multicolumn{4}{@{}l}{\textbf{Table\ref{tab:codebook}} (continued)} \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
Annotation & Behavior & Description & Example \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\endhead

@@

\subsection{Data preprocessing}
\label{subsec:datapreprocessing}
Utterances in MI session transcripts were segmented into successful and unsuccessful communication sequences. For each MI transcript, the stream of behavior codes from beginning to end was analyzed. Successful sequences were defined as those that resulted in a patient change talk or commitment language statement. Unsuccessful sequences were similarly created for sequences resulting in sustain talk. A total of 1,360 sequences were generated using this approach. The majority of the sequences (n=1,102) were successful, which is expected for a treatment-seeking population, in which the initial motivation for behavior change is typically high. Successful sequences had an average length of 5.28 utterances, while unsuccessful sequences had on average 5.29 utterances.

\subsection{Data Modeling}
\label{subsec:dataModeling}
\subsubsection{Hidden Markov Model}
\label{subsubsec:hiddenMarkovModel}
We applied the Hidden Markov Model (HMM)\footnote{we used the implementation in the hmmlearn package publicly available at http://hmmlearn.readthedocs.io/} to identify clusters of behavior codes corresponding to successful and unsuccessful communication sequences and to describe the relationships (transitions) between these clusters. Given a set of behavior code sequences, the posterior inference of HMM parameters involves the deduction of a temporal sequence of hidden states that best explains observations in each sequence. The rows in the emission probability matrix correspond to the distribution of observation symbols (i.e., the MYSCOPE behaviors displayed) for each hidden state and the transition probability matrix describes the transitions between the hidden states. Training an HMM with a given number of hidden states ($N$) involves estimating the following parameters using the Baum-Welch algorithm:

\begin{itemize}
\item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
\item $T$ is an $NXN$ state transition probability matrix, in which $t_{ij}$ is the probability of HMM transitioning from state $i$ to state $j$  
\item $E$ is an $NXM$ emission probability matrix, in which $e_{jk}$ is the probability of observing symbol $k$, when HMM is in state $j$ 
\item $p$ is the initial state distribution vector where $\pi_i$ is the probability of the first HMM state
\end{itemize}

We trained two HMM models, one using all successful sequences and the other one using all unsuccessful sequences. Each model was trained with the objective of maximizing the log-likelihood of all observations in the corresponding set of sequences. The optimal number of hidden states was determined by estimating the Bayesian information criterion (BIC) of HMM models with a different number of hidden states and selecting the model with the smallest value of BIC, which takes into account both log-likelihood and a penalty term for the number of parameters in the model to avoid overfitting. Experiments with a different number of hidden states in HMMs estimated on successful and unsuccessful sequences indicated that 5 hidden states were optimal for successful sequences and 2 hidden states were optimal for unsuccessful sequences (Figure~\ref{fig:bic}).

>>>>

\usepackage{rotating}
\usepackage{tabularx}
\usepackage{longtable,tabu}
\usepackage[flushleft]{threeparttable}
\usepackage{caption}
\captionsetup{justification=raggedright,singlelinecheck=off,labelfont=bf}
\usepackage{color, colortbl}
\usepackage{soul}
\definecolor{lightgraycol}{gray}{0.95}

@@

\title{\hl{Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis}
}

@@

\begin{abstract}
\hl{Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change. Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research. A significant barrier to this research is the time and resource-intensive cognitive tasks, including the manual coding and sequential analysis of coded transcripts, that are an integral part of the qualitative analysis of patient-counselor communication. Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating these cognitive tasks. In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had been previously annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients' motivational statements to guide clinical practice.}

\keywords{hidden markov model \and closed frequent pattern mining \and motivational interviewing \and \hl{pediatric obesity}}
\end{abstract}

\section{Introduction}
\label{sec:intro}
\hl{Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change}~\cite{miller2013motivational}. \hl{The theory underlying MI's clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills}~\cite{miller2013motivational}. \hl{The relational hypothesis suggests that counselors' use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change. MI's technical hypothesis}~\cite{miller2009toward} \hl{states that counselors' use of communication techniques consistent with the MI framework (``MI-consistent'' or MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''. Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change. Previous studies~}\cite{apodaca2009mechanisms} \hl{have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention}~\cite{walker2011influence}. \hl{In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk}~\cite{magill2014technical}. \hl{These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers' ability to test MI's technical hypothesis.}

\hl{Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations}~\cite{bakeman1997observing, bakeman2011sequential}. \hl{Moyers and Martin}~\cite{moyers2006therapist} \hl{applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN. A follow up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk}~\cite{moyers2009session}. \hl{Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk. Gaume et al.}~\cite{gaume2008counsellor} \hl{used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service. They found MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed. A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk}~\cite{gaume2010counselor}. \hl{In this sample, MIIN led to greater sustain talk, but was unrelated to change talk. Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk. Glynn and colleagues}~\cite{glynn2014change} \hl{linked reflections of change talk to the elicitation of change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use. In a study of adolescents engaged in weight loss treatment, Carcone et al.}~\cite{carcone2013provider} \hl{used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy. A parallel study of the adolescents' caregivers}~\cite{jacques2016building} \hl{drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk. Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in which the various MICO strategies are effective.}

\hl{The sequential analysis procedure used in the above MI process studies}~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} \hl{is based on the first-order Markov Chain model}~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}. \hl{The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event. Hence, Markov chain models have two main drawbacks. The first is their inability to preserve the long-range dependencies between observations in a sequence. However, in MI, an observed behavior can be influenced by any of the preceding behaviors. The second drawback is their inability to consider similarities between behavior codes and, consequently, their inability to identify multiple similar behaviors that lead to the same outcome. Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences. Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-term dependencies between behaviors, to identify causal relationships. The goal of the current research is to test the applicability of pattern data mining and machine-learning methods to identify effective patterns of patient-counselor communication.}

\hl{Several prior works have reported the results of adopting machine learning methods, such as topic models}~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, \hl{classification methods}~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} \hl{and neural networks}~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} \hl{to the tasks of annotating MI transcripts and assessing intervention fidelity. Perez-Rosas et al.}~\cite{perez2017predicting} \hl{developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews. Because MI fidelity is assessed in part by the ratio of counselor questions to reflection (a minimum goal is 1:1), their system employed a SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features achieving nearly 90\% accuracy. In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes}~\cite{hasan2016study}. \hl{The SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of MI transcripts with 17 behavior codes, which is comparable to human coders. In a follow up study}~\cite{hasan2018predicting}, \hl{we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk). RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews. This paper represents the next phase in this line of research. We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.}

\hl{HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden state'' corresponds to a different distribution. Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state. HMMs were originally proposed for speech recognition}~\cite{rabiner1989tutorial}, \hl{in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling}~\cite{uuguz2007biomedical, amrhein2003client}. \hl{For example, a HMM-based classifier was applied Doppler ultrasound imaging data to extract features from from the images that were then used to distinguish healthy patients from those with heart disease}~\cite{uuguz2007biomedical}. \hl{In another study, HMM was used to capture important
characteristics of protein families}~\cite{amrhein2003client}. \hl{In the application of HMM to patient-counselor communication, hidden states, and the sets of related behavior codes associated with the hidden states, may correspond a patient’s underlying motivational state during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans, and action steps toward behavior change}~\cite{baldi1994hidden}) \hl{as the antecedents of patients' behavior change}~\cite{apodaca2009mechanisms}, \hl{there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that are likely to lead to patient change talk and commitment language.}

\hl{Frequent pattern mining}~\cite{aggarwal2014frequent} \hl{is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together. Agrawal and Srikant}~\cite{agrawal1994fast} \hl{first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases. Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics}~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, \hl{medical imaging}~\cite{olukunle2002fast}, \hl{chemical and biological analysis}~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, \hl{web mining}~\cite{srivastava2000web}, \hl{and outlier analysis}~\cite{aggarwal2015outlier}. \hl{No published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication. A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns. To address this problem, we utilized the closed frequent itemset mining method}~\cite{pasquier1999discovering}, \hl{which produces fewer patterns in more compact form that are easier to interpret. In this study, we leveraged FPClose}~\cite{grahne2005fast}, \hl{an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk. FPClose is the best closed frequent itemset mining algorithm, which has shown highest performance in terms of running time and memory consumption.}

\hl{In this paper, we focus on computational methods to facilitate the sequential analysis of MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions. Analysis of these patterns will provide empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge will inform MI theory by providing additional evidence to support MI's technical hypothesis. It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication. This paper is the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions. Bertholet et al.}~\cite{bertholet2010change} \hl{used HMM to identify hidden states in a brief motivational intervention. Namely, they limited their HMM model to three hidden states which they characterized as ``towards change'', ``away from change'' and ``non-determined''. These states were then used as predictors in a bivariate regression model to predict drinking outcomes 12 months post-intervention. In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication. The goal of this study to evaluate the utility of using HMM and frequent pattern mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions. These two approaches offer the following advantages over the first-order Markov chain based methods most typically used in MI research. First-order Markov chain models identify the likely transitions between individual behaviors. In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications. Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors. Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communications during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.}

\section{\hl{Materials and Methods}}
\label{sec:materialsAndMethods}

\subsection{Dataset}
\label{subsec:dataset}
\hl{The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)}~\cite{carcone2013provider}. \hl{The MYSCOPE is an adaptation of the original MI-SCOPE}~\cite{martin2005motivational}, \hl{a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)}~\cite{moyers2005assessing}, \hl{the MI Skill Code (MISC)}~\cite{apodaca2007motivational}, \hl{and Amrhein's conceptualization of change talk and commitment}~\cite{amrhein2004does}. \hl{A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table}~\ref{tab:codebook}.

\begin{small}
\begin{longtable}{p{1.3cm}p{2cm}p{5cm}p{4.8cm}}
\caption{\hl{MYSCOPE codebook}} 
\label{tab:codebook} \\
\hline\noalign{\smallskip}
Annotation & Behavior & Description & Example \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
\endfirsthead
\multicolumn{4}{@{}l}{\textbf{Table\ref{tab:codebook}} (continued)} \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
Annotation & Behavior & Description & Example \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\endhead

@@

\subsection{Data preprocessing}
\label{subsec:datapreprocessing}
\hl{Utterances in MI session transcripts were segmented into successful and unsuccessful communication sequences. For each MI transcript, the stream of behavior codes from beginning to end was analyzed. Successful sequences were defined as those that resulted in a patient change talk or commitment language statement. Unsuccessful sequences were similarly created for sequences resulting in sustain talk. A total of 1,360 sequences were generated using this approach. The majority of the sequences (n=1,102) were successful, which is expected for a treatment-seeking population, in which the initial motivation for behavior change is typically high. Successful sequences had an average length of 5.28 utterances, while unsuccessful sequences had on average 5.29 utterances.}

\subsection{Data Modeling}
\label{subsec:dataModeling}
\subsubsection{Hidden Markov Model}
\label{subsubsec:hiddenMarkovModel}
\hl{We applied the Hidden Markov Model (HMM)\footnote{we used the implementation in the hmmlearn package publicly available at http://hmmlearn.readthedocs.io/} to identify clusters of behavior codes corresponding to successful and unsuccessful communication sequences and to describe the relationships (transitions) between these clusters. Given a set of behavior code sequences, the posterior inference of HMM parameters involves the deduction of a temporal sequence of hidden states that best explains observations in each sequence. The rows in the emission probability matrix correspond to the distribution of observation symbols (i.e., the MYSCOPE behaviors displayed) for each hidden state and the transition probability matrix describes the transitions between the hidden states. Training an HMM with a given number of hidden states ($N$) involves estimating the following parameters using the Baum-Welch algorithm:}

\begin{itemize}
\item \hl{$M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table}~\ref{tab:codebook})
\item \hl{$T$ is an $NXN$ state transition probability matrix, in which $t_{ij}$ is the probability of HMM transitioning from state $i$ to state $j$} 
\item \hl{$E$ is an $NXM$ emission probability matrix, in which $e_{jk}$ is the probability of observing symbol $k$, when HMM is in state $j$ }
\item \hl{$p$ is the initial state distribution vector where $\pi_i$ is the probability of the first HMM state}
\end{itemize}

\hl{We trained two HMM models, one using all successful sequences and the other one using all unsuccessful sequences. Each model was trained with the objective of maximizing the log-likelihood of all observations in the corresponding set of sequences. The optimal number of hidden states was determined by estimating the Bayesian information criterion (BIC) of HMM models with a different number of hidden states and selecting the model with the smallest value of BIC, which takes into account both log-likelihood and a penalty term for the number of parameters in the model to avoid overfitting. Experiments with a different number of hidden states in HMMs estimated on successful and unsuccessful sequences indicated that 5 hidden states were optimal for successful sequences and 2 hidden states were optimal for unsuccessful sequences (Figure}~\ref{fig:bic}).
